---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r LoadData, echo = T}
library(data.table)
DT <- fread('activity.csv', colClasses = c('integer', 'character', 'integer'))
DT[, date := as.Date(date)]
```

## What is mean total number of steps taken per day?


```{r, echo = T, fig.width= 9, fig.height=5}
steps.each.day <- DT[!is.na(steps), .(NSteps=sum(steps)), by = date]
library(ggplot2)
# plot histogram
ggplot(steps.each.day, aes(NSteps)) + 
  geom_histogram(binwidth=1000, colour = 'white', fill = 'lightblue')+
  labs(x='Total number of steps per day', 
       title = 'Histogram of total steps per day (NAs are excluded)')
```

```{r, echo = T, results = 'asis'}
library(pander)
pandoc.table(summary(steps.each.day$NSteps),
             style = 'rmarkdown', 
             caption = 'Summary of total steps each day')
```


## What is the average daily activity pattern?

```{r, echo=T}
avg.steps.5m <- DT[, .(avg=mean(steps, na.rm = T)), by = interval]
qplot(data = avg.steps.5m, interval, avg, geom='line')+
  geom_line(colour = 'darkgreen')+ 
  labs(x = 'Time, minute', y = 'Mean num. steps',
       title = 'Average step counts per 5-minute (NAs excluded)')
```

```{r, echo=T}
# Order by average steps (descending)
avg.steps.5m[order(-avg), ]
```

## Imputing missing values
```{r echo=T}
num.NA <- sum(is.na(DT$steps))
num.NA
```
Unfortunately, there are `r num.NA` records
with NA in the dataset. Here we assign the average value 
to missing fields.  Missing step counts (i.e., NAs) are
repalced by the average count corresponding to
the interval.  

```{r, echo=T}
setkey(avg.steps.5m, interval)
missing.set <- DT[is.na(steps), ]
setkey(missing.set, interval)
# Look up mean steps for a interval and fill in missing set
temp <- avg.steps.5m[missing.set, .(avg, interval, date)]
setnames(temp, c('steps', 'interval', 'date'))
x <- DT[!is.na(steps), ]
new.DT <- rbind(x, temp)
```


```{r, echo = T, fig.width= 9, fig.height=5}
steps.each.day <- new.DT[, .(NSteps=sum(steps)), by = date]
ggplot(steps.each.day, aes(NSteps)) + 
  geom_histogram(binwidth = 1000,
                 colour = 'white', fill = 'lightblue')+
  labs(x='Total number of steps per day', 
       title='Updated dataset with NA replaced by mean step count')
```

For the updated dataset, let's look at some statistics of steps per day.

```{r, echo = T, results = 'asis'}
pandoc.table(summary(steps.each.day$NSteps), style = 'rmarkdown')
```

## Are there differences in activity patterns between weekdays and weekends?


```{r, echo = T}
new.DT[, ':='(weekday=weekdays(date),
              group='weekday')]
new.DT[weekday=='Sunday'|weekday=='Saturday', group:='weekend']
new.DT$group <- factor(new.DT$group, levels = c('weekday','weekend'))
avg.step <- new.DT[group=='weekday', .(avg=mean(steps)), by=interval]
avg.step.2 <- new.DT[group=='weekend', .(avg=mean(steps)), by=interval]
plot.df <- merge(x = avg.step, y = avg.step.2,
                 by = 'interval', all = T)
setnames(plot.df, c('interval', 'weekday', 'weekend'))
library(reshape2)
plot.df.m <- melt(plot.df, id.vars = 'interval',
                  variable.name = 'group', value.name = 'avg')
ggplot(plot.df.m, aes(x = interval, y = avg)) +
  geom_line() + facet_grid(group ~ .) + 
  labs(x = 'Time, minute', y = 'Average steps per 5 minutes',
       title = 'Weekday/Weekend activity pattern comparison\nNAs are replaced with mean num. of steps')
```
